MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-21 Manuscript/Manuscript-1-DCA.Rmd
vsn 1 edits

For many macroevolutionary analyses, it is often of interest to quantify the degree to which phylogenetic signal is displayed in continuous traits. Several analytical tools have been developed for this purpose [e.g., @Gittleman1990; @Abouheif1999; @Pagel1999; @Blomberg_et_al2003; @Klingenberg2010; @Adams2014a], which differ primarily in how they characterize the phylogenetic dependency of trait variation among taxa. One commonly used statistical measure, *Kappa* ($K$), expresses the strength of phylogenetic signal as the ratio of observed trait variation to the trait variation conditioned on the phylogeny; scaled by what is expected under Brownian motion given the phylogeny's size and shape [@Blomberg_et_al2003; also @Adams2014a]. Another approach, Pagel's $\lambda$ [@Pagel1999], uses maximum likelihood to fit the data to the phylogeny under some model of evolutionary change (typically Brownian motion). The inclusion of a scaling parameter, $\lambda$,  transforms the lengths of the internal branches of the phylogeny to improve the fit, and this parameter describes the degree of phylogenetic signal in the dataset [@Pagel1999; @Freckleton_et_al2002]. Pagel's $\lambda$ also has the advantage that it may be included when estimating the association of traits in a phylogenetic context, meaning that one may account for the degree of phylogenetic signal while conducting phylogenetic regression or ANOVA [see @Freckleton_et_al2002]. \hfill\break

Several studies have investigated the statistical properties of methods for estimating phylogenetic signal under various conditions [e.g., @Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @MolinaVenegas2017; see also @Revell_et_al2008; @Revell2010]. These have largely focused on the ability of methods to detect the presence of phylogenetic signal (i.e., type I and type II error rates) under complex models of evolutionary change, across a range of phylogeny sizes, and with varying degrees of phylogenetic uncertainty or unresolved topologies. In terms of parameter estimation, Revell [-@Revell2010] found that regression parameters were accurately estimated when $\lambda$ was included during phylogenetic regression, and Munkmuller et al. [-@Munkemuller_et_al2012] demonstrated that estimates of phylogenetic signal obtained using various measures generally increased when input levels of phylogenetic signal were stronger. However, the precision of those estimates could not be determined, because the input levels of phylogenetic signal were simulated via a scaling factor ($w$) that was not directly comparable to the measures of phylogenetic signal being compared [see @Munkemuller_et_al2012]. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$), concluding that insufficient data (i.e., the number of species) was the underlying cause of the lack of precision in parameter estimation. However, this conclusion assumes that the precision in estimating $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. First we use computer simulations across differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. For comparison, we also evaluate estimates of $Kappa$ and its empirically-derived effect size ($Z_K$) for this purpose. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic regression and ANOVA (i.e., PGLS) affects parameter estimation, and whether levels of phylogenetic signal estimated in a PGLS framework are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. Further, we find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that this variation is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. By contrast, variation across effect sizes ($Z_K$) obtained from $Kappa$ is far more consistent across the range of input values, and the precision of estimating $Z_K$ is preferable to that observed using $\lambda$. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often innacurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure should be treated with caution. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and comparing the strength of phylogenetic signal across datasets.

@@

Using the simulated phylogenies we estimated the phylogenetic signal in each dataset using Pagel's $\lambda$. Next, for the 1050 datasets at each level of species richness ($n$), we assessed the relationship between $\lambda_{in}$ and $\lambda_{est}$ using regression. Additionally, the precision of $\lambda$ was approximated by observing the variation of $\lambda_{est}$ obtained at each input level of phylogenetic signal ($\lambda_{in}$). Finally, for comparison we estimated the degree of phylogenetic signal in each dataset using $Kappa$ [@Blomberg_et_al2003; @Adams2014a], and from this calculated its effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] for each dataset. Here, the effect size was estimated as: $Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}$, where $K_{obs}$ was the observed $Kappa$, and $\mu_K$ and $\sigma_K$ were the mean and standard deviation of the empirical sampling distribution of values obtained from the permutation distribution [for discussion see: @AdamsCollyer2019]. Because $Z_K$ differs in scale from $\lambda$, we used a linear normalization to standardize $Z_K$ to a uniform distribution ($0\rightarrow1$), and variation in the set of normalized $Z_K$ values at each input level of phylogenetic signal ($\lambda_{in}$) were then estimated for comparison. \hfill\break

@@

To determine how Pagel's $\lambda$ is commonly utilized in empirical studies, we conducted a literature survey. From Google.scholar we obtained a list of all papers published in 2019 that used $\lambda$; resulting in 341 studies. For each study, we extracted all $\lambda_{est}$, the size of the phylogeny ($n$), and noted whether authors reported confidence intervals or performed significance tests assessing difference of $\lambda_{est}$ from either zero or one. We also noted whether biological interpretations based on $\lambda_{est}$ were made, and for studies that reported more than one $\lambda_{est}$, we also noted whether these were compared in some manner, and whether such comparisons were accompanied with statistical tests between $\lambda_{est}$. \hfill\break

@@

Our primary series of simulations revealed several patterns. First, the relationship between $\lambda_{est}$ and input levels of phylogenetic signal was $\sim\beta=1.0$ at each level of species richness ($n=32, 64, 128, 256, 512, 1024$). This implied that the mean $\lambda_{est}$ across datasets for a given set of simulation conditions reflected input levels of phylogenetic signal. However, as shown in Figure 1, the precision of those estimates varied widely, and in several interesting ways. Predictably, precision was worst at low levels of species richness, where in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{est}=0.5$: range = $0.0\to0.985$). Second, as species richness increased, variation across estimates of $\lambda$ decreased (Figure 1). This confirmed the pattern identified by Boettiger et al. [-@Boettiger_et_al2012] on  a small ($n=13$) and large ($n=281$) phylogeny, and demonstrated that this trend was general across a wider range of species richness. Importantly however, our broader set of simulations revealed that the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal, and that there was a particular pattern to the imprecision of estimating this phylogenetic parameter. Specifically, we found that variation in $\lambda_{est}$ was highest at intermediate levels of phylogenetic signal, and decreased at both lower and higher levels of input signal (Figure 1). This implied that the precision in estimating $\lambda$ was worst at intermediate levels of phylogenetic signal; a pattern that remained consistent across all levels of species richness evaluated here (Figure 1). Notably, even at very large levels of species richness, the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=1024$; $\lambda_{est}=0.5$: range = $0.31\to 0.68$). From this we conclude that $\lambda_{est}$ is not an accurate estimator of the strength of phylogenetic signal in phenotypic datasets, and that biological interpretations of the level of phylogenetic signal based on $\lambda$ should be treated with caution. \hfill\break

@@

**STOPPED HERE**By contrast,  (FIg 2) Kappa. Much smaller range and variation as compared with lambda, but with larger l-in, variation in kappa does increase. However, this is not the case with Z-K.  Z-K increases predictably (linearly, progressively...) with L-in, and importantly variation in Z-K remains nearly constant across the range of input values. this means that Z-K is equally precise in estimating the strength of phylogenetic signal, and may be more adept at facilitating useful statements on teh relative strength of phylogenetic signal within and across datasets. \hfill\break

**NOTE: Fig 2 is 4 panels for N=128: Lambda, Kappa, Z-K, Z-k-normalized.**

@@

@@

[insert Figure 3 here] \hfill\break

>>>>

For many macroevolutionary analyses, it is often of interest to quantify the degree to which phylogenetic signal is displayed in continuous traits. Several analytical tools have been developed for this purpose [e.g., @Gittleman1990; @Abouheif1999; @Pagel1999; @Blomberg_et_al2003; @Klingenberg2010; @Adams2014a], which differ primarily in how they characterize the phylogenetic dependency of trait variation among taxa. One commonly used statistical measure, *Kappa* ($K$), expresses the strength of phylogenetic signal as the ratio of observed trait variation to the trait variation conditioned on the phylogeny; scaled by what is expected under Brownian motion given the phylogeny's size and shape [@Blomberg_et_al2003; also @Adams2014a]. Another approach, Pagel's $\lambda$ [@Pagel1999], uses maximum likelihood to fit the data to the phylogeny under some model of evolutionary change (typically Brownian motion). The inclusion of a scaling parameter, $\lambda$,  transforms the lengths of the internal branches of the phylogeny to improve the fit, and this parameter describes the degree of phylogenetic signal in the dataset [@Pagel1999; @Freckleton_et_al2002]. Pagel's $\lambda$ also has the appeal that it may be included when estimating the association of traits in a phylogenetic context, meaning that one may account for the degree of phylogenetic signal while conducting phylogenetic regression or ANOVA [see @Freckleton_et_al2002]. \hfill\break

Several studies have investigated the statistical properties of methods for estimating phylogenetic signal under various conditions [e.g., @Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @MolinaVenegas2017; see also @Revell_et_al2008; @Revell2010]. These have largely focused on the ability of methods to detect the presence of phylogenetic signal (i.e., type I and type II error rates) under complex models of evolutionary change, across a range of phylogeny sizes, and with varying degrees of phylogenetic uncertainty or unresolved topologies. In terms of parameter estimation, Revell [-@Revell2010] found that regression parameters were accurately estimated when $\lambda$ was included during phylogenetic regression, and Munkmuller et al. [-@Munkemuller_et_al2012] demonstrated that estimates of phylogenetic signal obtained using various measures generally increased when input levels of phylogenetic signal were stronger. However, the precision of those estimates could not be determined, because the input levels of phylogenetic signal were simulated via a scaling factor ($w$) that was not directly comparable to the measures of phylogenetic signal being compared [see @Munkemuller_et_al2012]. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$), concluding that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates. However, this conclusion assumes that the precision in estimating $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We also evaluate the precision of a standardized effect size ($Z_K$) based on $Kappa$ for comparison. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic regression and ANOVA (i.e., PGLS) affects parameter estimation, and whether levels of phylogenetic signal estimated in a PGLS framework are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. We also find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that this variation is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. In addition, the same $\lambda_{est}$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. By contrast, variation across effect sizes ($Z_K$) obtained from $Kappa$ is far more consistent across the range of input values, and the precision of $Z_K$ is generally preferable to that of $\lambda_{est}$. Further, because $Z_K$ is a standardized effect size, the strength of phylogenetic signal is more readily interpretable, and may be compared across datasets. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often innacurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

@@

Using the simulated phylogenies we estimated the phylogenetic signal in each dataset using Pagel's $\lambda$. Next, for the 1050 datasets at each level of species richness ($n$), we assessed the relationship between $\lambda_{in}$ and $\lambda_{est}$ using regression. Additionally, the precision of $\lambda$ was approximated by observing the variation of $\lambda_{est}$ obtained at each input level of phylogenetic signal ($\lambda_{in}$). Finally, for comparison we characterized the strength of phylogenetic signal in each dataset using a standardized effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] based on $Kappa$. As suggested by Adams and Collyer [-@AdamsCollyer2019], the effect size for phylogenetic signal may be estimated as: $Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}$, where $K_{obs}$ was the observed $Kappa$, and $\mu_K$ and $\sigma_K$ were the mean and standard deviation of the empirical sampling distribution of values obtained from the permutation distribution. $Z_K$ describes the strength of phylogenetic signal as a standard deviate from its sampling distribution, and thus directly measures the strength of signal in a manner that is comparable across datasets. Variation in the set of $Z_K$ at each input level of phylogenetic signal was then calculated as an estimate of precision in $Z_K$. However, because $Z_K$ differs in scale from $\lambda$, we used a linear normalization to standardize $Z_K$ to a uniform distribution ($0\rightarrow1$), and estimated the precision of $Z_K$ from the normalized values. \hfill\break

@@

To determine how Pagel's $\lambda$ is commonly utilized in empirical studies, we conducted a literature survey. From Google.scholar we obtained a list of all papers published in 2019 that used $\lambda$; resulting in 341 studies. For each study, we extracted all $\lambda_{est}$, the size of the phylogeny ($n$), and noted whether authors reported confidence intervals or performed significance tests assessing difference of $\lambda_{est}$ from either zero or one. We also noted whether biological interpretations based on $\lambda_{est}$ were made, and for studies that reported more than one $\lambda_{est}$, we also noted whether these were compared in some manner, and whether such comparisons were accompanied with statistical tests between $\lambda_{est}$.

@@

Our first set of simulations revealed several patterns. First, the relationship between $\lambda_{est}$ and $\lambda_{in}$ across the range input levels of phylogenetic signal was $\sim\beta=1.0$; indicating that the average $\lambda_{est}$ across datasets for a given set of simulation condition correctly reflected input levels of phylogenetic signal. However, as shown in Figure 1, the precision of those estimates varied widely, and in several interesting ways. Predictably, precision was worse at low levels of species richness, where in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: range of $\lambda_{est}= 0.0\to 0.985$). Second, as species richness increased, variation across estimates of $\lambda$ decreased (Figure 1). This confirmed the pattern identified by Boettiger et al. [-@Boettiger_et_al2012] on  a small ($n=13$) versus a large ($n=281$) phylogeny, and demonstrated that the trend of increasing precision with higher species richness was general. Importantly however, our broader set of simulations revealed that the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal. Specifically, we found that variation in $\lambda_{est}$ was highest at intermediate levels of phylogenetic signal, and decreased at both lower and higher levels of input signal (Figure 1). This implied that the precision in estimating $\lambda$ was worse at intermediate values, and improved as the levels of phylogenetic signal were closer to $\lambda_{in}=0$ (no phylogenetic signal) or $\lambda_{in}=1$ (Brownian motion). Notably, even at large levels of species richness, the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: range of $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; range of $\lambda_{in} = 0.25 \to 0.65$). Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize observed levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

By contrast, when characterizing phylogenetic signal with the standardized effect size ($Z_K$) of $Kappa$, we found that the precision of $Z_K$ was more stable, as variation across datasets was far more consistent across the range of input values. For example, when $n=128$ the precision of $\lambda_{est}$ (Figure 2A) varied considerably more across input levels of phylogenetic signal than did the precision of $Z_K$ for the same datasets (Figure 2B). Further, for a given $n$, the variance in precision estimates of $Z_K$ was considerably smaller than the variance in precision estimates of $\lambda_{est}$ (Fig. 2C,D); implying that estimates of phylogenetic signal were more reliable and robust when using $Z_K$ (for additional results see Supporting Information).  Finally, it should also be recognized that because $Z_K$ is a standardized effect size, the strength of phylogenetic signal is more readily interpretable when using this measure, as $Z_K$ expresses the strength of phylogenetic signal in standard deviation units relative to the mean [see @AdamsCollyer2019]. This further implies that comparisons of the strength of phylogenetic signal among phenotypic traits are possible, and may be accomplished statistically via a two-sample test that formally compares $Z_K$ across datasets [for comparisons of multivariate effect sizes see: @AdamsCollyer2016; @AdamsCollyer2019b]. \hfill\break

@@

When $\lambda$ was incorporated in PGLS regression and ANOVA (i.e., $Y\sim{X}$), we found much the same pattern as in the previous simulations. Namely, the precision of $\lambda_{est}$ varied inversely with species richness; where greater precision was obtained at higher levels of species richness (Figure 3). Likewise, the precision in estimating $\lambda$ was worse at intermediate values (Figure 3), and improved as the levels of phylogenetic signal were closer to $\lambda_{in}=0$ (no phylogenetic signal) or $\lambda_{in}=1$ (Brownian motion). Generally, the spread of estimates was slightly broader when $\lambda$  was co-estimated with regression parameters in PGLS regression (Figure 3), as compared to when $\lambda$ was estimated in cases only the dependent variable (Y) was considered (Figure 1). Finally, found that regression parameters ($\beta$) were accurately estimated when $\lambda$ was included during phylogenetic regression; a result which confirmed earlier findings of Revell [-@Revell2010] (see Supporting Information). \hfill\break

[insert Figure 3 here]

@@

[insert Figure 4 here] \hfill\break
